\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-16 NAR_DRAFT.tex
fixing rest of figure calls

\begin{abstract}
%The accumulation of somatic mutations is an inevitable consequence of tumor development.
Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in treatment resistance, and the value of tumor heterogeneity as a diagnostic biomarker.
The ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
% \rngcomment{A little too much intro in abstract.}
% This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
Here we present \batcave, an algorithm that uses variant calls produced by MuTect to compute a tumor and site-specific prior probability of mutation for every potential variant, and uses this prior probability to calibrate the posterior probability for every variant.
\batcave uses the mutation profile and allele frequency spectrum of high-confidence mutations to estimate the data generating distribution giving rise to variants.
We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
The algorithm performs consistently as sequencing depth increases, enabling deeper sequencing and the identification lower frequency variants without generating excessive false positive calls.
\rngcomment{Need a sentence about what new analysis this enables. How does \batcave (at least help) solve the problem you raised earlier?}
We provide an R package implementing the \batcave algorithm for use with MuTect output.
\end{abstract}

@@

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, Figure \ref{NAR-sigfig}\textbf{A} through \textbf{C}, shows mutation profiles from several tumors.
\rngcomment{Check how NAR Genomics and Bioinformatics formats figure legends, equation references, etc.}
\bkmcomment{Fixed in portions of text. need to change all figures to have capital letter panel labels. A and so on.}
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/signatures_only.pdf}
  \end{center}
  \caption{Tumor mutation profiles.
  In each panel, the y-axis is the proportion of total mutations in the tumor with each of the 96 substitution types shown on the x-axis.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor we used in our real data analysis \cite{Shi2018}.
  \textbf{(C)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(D)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(E)} Equal proportions of signatures 1, 3, \& 5.
  \textbf{(F)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al. (2019) \citep{Alexandrov2019}.
  \textbf{(G)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al. (2019) \citep{Alexandrov2019}.
  }
  \label{NAR-sigfig}
 \end{figure}

@@

  \begin{figure*}
    \begin{center}
    \includegraphics{figures/fig_wes.pdf}
    \end{center}
    \caption{Model performance.
    \textbf{(a)} Precision-recall curves and \textbf{(b)} Receiver operating characteristic curves for 3 mutation signatures.
    Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{(c)} the MuTect model and \textbf{(d))} the \batcave model.
    A perfectly calibrated model would generate the diagonal line. Precision-recall curves for real data \textbf{(e)} and \textbf{(f)}.}
  \label{NAR-wes_fig}
  \end{figure*}

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/kl_only.pdf}
  \end{center}
  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(f)} Whole exomes and \textbf{(g)} whole genomes.
  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
\label{NAR-kl_fig}
\end{figure}

\subsection{Convergence of the site-specific prior to the data generating distribution}
The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
In Figure \ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the \citet{Griffith2015} leukemia is over 17000, and for our simulated data sets ranges between 1500 and 5000.

\subsection{Simulated data}
We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}.
These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.

We assessed classification performance with both the area under the ROC curve and the area under the precision-recall curve, because the classes here are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
By both metrics \batcave outperforms MuTect (Figure \ref{NAR-wes_fig} \textbf{a} and \textbf{c}, and Table \ref{table:01}).
The extent to which \batcave outperforms is dependent on both the sequencing depth and the level of concentration of the mutation signature.
Deeper sequencing and more concentrated mutation signatures increase the performance advantage of \batcave.

The estimated mutation rate is approximately $3\cdot10^{-7}$ for all simulated tumors (Table \ref{table:01}).
We note that the effective mutation rate is lower than the simulated rate of $3\cdot10^{-6}$.
In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.

We measured overall calibration performance using the Integrated Calibration Index (ICI) described in Methods \citep{Austin2019}.
The ICI is a measure of the difference between predicted and observed probabilities that is weighted by the empirical density of the predicted probabilities.
This is useful in our setting because the density of posterior probabilities tends to be strongly bi-modal.
A large fraction of true negative variants have posterior probabilities of less than $10{-4}$, which is a probability far below any threshold that is meaningful, we evaluate calibration only on variants with posterior probability greater than .01.
As we can see in Figure \ref{NAR-wes_fig} \textbf{b} and \textbf{d}, \batcave tends to increase posterior probabilities of low probability but true variants (density curves) while decreasing the probability of low probability but false variants (density not shown).
For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse (Figure \ref{NAR-wgs_fig} and Table \ref{table:01}).
This effect is likely due to the fact that there are simply very few low probability true positive variants in tumor sequenced to 100X depth.
As with all other metrics, the benefit of \batcave increases as the concentration of the mutation profile increases, and as sequencing depth increases.

Figure \ref{NAR-ppv_fig} takes a consolidated view of classification accuracy and calibration for 500X whole exomes in the portion of the posterior probability distribution relevant to clinicians and researchers.
For any probability between 60 and 90\% \batcave is better calibrated than Mutect.
Essentially, for any posterior probability above 70\% MuTect has a false positive rate around 8\%, while \batcave has a false positive rate that increases as posterior probability decreases.
The cost of this compression is recall, which is 3-5\% better than MuTect at 95\% posterior probability, and 3-7\% better at 90\%, depending on muation profile concentration.
At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect.
Figure \ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
Note that recall never improves because the uncalled true positive variants are below the threshold of detection for any reasonable level of false positive calls.

\subsection{Real tumor data.}
We utilized two real data sets.
In each, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
These are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
\citet{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{a} and \textbf{b} and Table \ref{table:01}).

\citet{Shi2018} performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$, and validated all variants identified by three different variant calling pipelines.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure \ref{NAR-wes_fig} \textbf{b} and Table \ref{table:01}).

>>>>

\begin{abstract}
%The accumulation of somatic mutations is an inevitable consequence of tumor development.
Most of the mutations in a tumor at time of diagnosis are present in only a small fraction of all tumor cells.
Identification of these low frequency mutations is increasingly important as evidence has accumulated for their role in treatment resistance, and the value of tumor heterogeneity as a diagnostic biomarker.
The ability to reliably detect low frequency mutations increases as sequencing depth increases, yet false positive identification rates also increase with sequencing depth.
% This dynamic has historically hindered the use of sequencing depths much greater than about 150X average coverage for non-targeted sequencing such as whole exome or whole genome.
Here we present \batcave, an algorithm that uses variant calls produced by MuTect to compute a tumor and site-specific prior probability of mutation for every potential variant, and uses this prior probability to calibrate the posterior probability for every variant.
\batcave uses the mutation profile and allele frequency spectrum of high-confidence mutations to estimate the data generating distribution giving rise to variants.
We show that high-confidence variant calls are sufficient to generate a good estimate of this data generating distribution, and the posterior probabilies of mutation for called variants are better calibrated than those arising from current variant calling algorithms.
The algorithm performs consistently as sequencing depth increases, enabling deeper sequencing and identification of lower frequency variants while providing well calibrated false positive rate expectations.
We provide an R package implementing the \batcave algorithm for use with MuTect output.
\end{abstract}

@@

The mutational processes that generate single nucleotide variants in tumors do not act uniformly across the genome.
%Single nucleotide variants arise in tumors at a rate and at genomic locations driven by two main processes. 
The processes of spontaneous mutation that are active in all somatic tissues depend sensitively on local nucleotide context \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}. 
Additional mutational processes are active in tumors, due to mutagen exposure or defects in DNA maintenance and repair, and these processes are also sensitive to local nucleotide context \citep{Alexandrov2013a,Helleday2014a,Nik-Zainal2016,Kandoth2013,Alexandrov2016}.
The specific mutational processes active in a particular tumor generate its unique mutation profile, and differences within and between tumor types are pronounced \cite{Stephens2005, Burrell2013a, Nakamura2015, Witkiewicz2015, Kumar2016}.
For example, Figure \ref{NAR-sigfig}\textbf{A}--\textbf{C}, shows mutation profiles from several tumors.
\rngcomment{Check how NAR Genomics and Bioinformatics formats figure legends, equation references, etc.}
\bkmcomment{Fixed in portions of text. need to change all figures to have capital letter panel labels. A and so on.}
%Figure \ref{NAR-sigfig} \textbf{a-e} shows several examples of mutation profiles from both real and simulated tumors.

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/signatures_only.pdf}
  \end{center}
  \caption{Tumor mutation profiles.
  In each panel, the y-axis is the proportion of total mutations in the tumor with each of the 96 substitution types shown on the x-axis.
  \textbf{(A)} The observed mutation profile of an acute myeloid leukemia used in our real data analysis \cite{Griffith2015}.
  \textbf{(B)} The observed mutation profile of a breast tumor we used in our real data analysis \cite{Shi2018}.
  \textbf{(C)} A mutation profile used for simulating tumors, made up of equal proportions of COSMIC mutation signatures 1, 7, \& 11.
  \textbf{(D)} Equal proportions of signatures 1, 4, \& 5.
  \textbf{(E)} Equal proportions of signatures 1, 3, \& 5.
  \textbf{(F)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al., 2019 \citep{Alexandrov2019}.
  \textbf{(G)} The observed mutaiton profile of a real breast tumor obtained from supplementary materials in Alexandrov et al., 2019 \citep{Alexandrov2019}.
  }
  \label{NAR-sigfig}
 \end{figure}

@@

  \begin{figure*}
    \begin{center}
    \includegraphics{figures/fig_wes.pdf}
    \end{center}
    \caption{Model performance.
    \textbf{(A)} Precision-recall curves and \textbf{(B)} Receiver operating characteristic curves for 3 mutation signatures.
    Distribution of estimated true positive probabilities for true positive (top) and true negative (bottom) variants for \textbf{(C)} the MuTect model and \textbf{(D))} the \batcave model.
    A perfectly calibrated model would generate the diagonal line. Precision-recall curves for real data \textbf{(E)} and \textbf{(F)}.}
  \label{NAR-wes_fig}
  \end{figure*}

@@

\begin{figure}
  \begin{center}
  \includegraphics{figures/kl_only.pdf}
  \end{center}
  \caption{KL divergence between the estimated profile ($\boldsymbol{\pi}$) and the simulated mutation profile for \textbf{(A)} Whole exomes and \textbf{(B)} whole genomes.
  The estimated mutation profile converges quickly to the target as the number of mutations evaluated rises.}
\label{NAR-kl_fig}
\end{figure}

\subsection{Convergence of the site-specific prior to the data generating distribution}
The context-dependent prior probability of mutation, the heart of \batcave, depends on the prior converging to a good representation of the data generating distribution within the set of high-confidence mutations.
We use the Kullback-Leibler (KL) divergence as a measure of the distance (or entropy difference) between the estimated prior and the actual simulated data generating distribution.
In Figure~\ref{NAR-kl_fig} we order mutations by descending MuTect probability value, and compute the KL divergence between the prior after adding each mutation and the data generating distribution.
For both whole exome and whole genome the prior converges to the data generating distribution easily within the number of available high-confidence mutations, which for the \citet{Griffith2015} leukemia is over 17000, and for our simulated data sets ranges between 1500 and 5000.

\subsection{Simulated data}
We generated six different tumor/normal pairs of two types, 3 each of 100X depth whole genomes and 500X whole exomes.
For each of the two types we simulated tumors with variants drawn from three different mutation profiles chosen to resemble melanoma (concentrated), a lung cancer (intermediate), and a BRCA-driven breast cancer (diffuse)  as described in Materials \& Methods and Figure \ref{NAR-sigfig}\textbf{C,D,E}.
These 6 simulated tumor/normal pairs were used to assess the effect of sequencing depth and mutation profile type on the results obtained from \batcave.

We assessed classification performance with both the area under the ROC curve and the area under the precision-recall curve, because the classes here are unbalanced (approximately 5 to 1 ratio of false to true variants in our simulated data).
By both metrics \batcave outperforms MuTect (Figure~\ref{NAR-wes_fig}\textbf{A}--\textbf{B}, and Table~\ref{table:01}).
The extent to which \batcave outperforms is dependent on both the sequencing depth and the level of concentration of the mutation signature.
Deeper sequencing and more concentrated mutation signatures increase the performance advantage of \batcave.

The estimated mutation rate is approximately $3\cdot10^{-7}$ for all simulated tumors (Table~\ref{table:01}).
We note that the effective mutation rate is lower than the simulated rate of $3\cdot10^{-6}$.
In this case the lower effective rate is not due to normal contamination of the tumor sample, but the fact that \texttt{Bamsurgeon} has restrictions -- such as sequencing depth and quality -- that prevent 100\% of simulated variants from being inserted.

We measured overall calibration performance using the Integrated Calibration Index (ICI) described in Methods \citep{Austin2019}.
The ICI is a measure of the difference between predicted and observed probabilities that is weighted by the empirical density of the predicted probabilities.
This is useful in our setting because the density of posterior probabilities tends to be strongly bi-modal.
A large fraction of true negative variants have posterior probabilities of less than $10^{-4}$, which is a probability far below any threshold that is meaningful, we evaluate calibration only on variants with posterior probability greater than .01.
As we can see from the comparison in Figure~\ref{NAR-wes_fig}\textbf{C}--\textbf{D}, \batcave tends to increase posterior probabilities of low probability but true variants (density curves) while decreasing the probability of low probability but false variants (density not shown).
For 500X tumors the calibration of \batcave is better across the full spectrum of posterior probabilities, while for 100X whole genomes the calibration is slightly worse (Figure \ref{NAR-wgs_fig} and Table \ref{table:01}).
This effect is likely due to the fact that there are simply very few low probability true positive variants in tumor sequenced to 100X depth.
As with all other metrics, the benefit of \batcave increases as the concentration of the mutation profile increases, and as sequencing depth increases.

Figure~\ref{NAR-ppv_fig} takes a consolidated view of classification accuracy and calibration for 500X whole exomes in the portion of the posterior probability distribution relevant to clinicians and researchers.
For any probability between 60 and 90\% \batcave is better calibrated than Mutect.
Essentially, for any posterior probability above 70\% MuTect has a false positive rate around 8\%, while \batcave has a false positive rate that increases as posterior probability decreases.
The cost of this compression is recall, which is 3-5\% better than MuTect at 95\% posterior probability, and 3-7\% better at 90\%, depending on muation profile concentration.
At any posterior probability \batcave returns a larger fraction of all positive calls than MuTect.
Figure~\ref{NAR-ppv_wgs_fig} shows results for 100X whole genomes, where the results are similar for precision.
Note that recall never improves because the uncalled true positive variants are below the threshold of detection for any reasonable level of false positive calls.

\subsection{Real tumor data.}
We utilized two real data sets.
In each, deep sequencing and variant validation were performed with the specific purpose of evaluating tumor variant calling pipelines.
These are among a very small number of tumors which have been sequenced at high depth, and for which variant validation was undertaken.
\citet{Griffith2015} sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>360\mathrm{X}$, and used targeted sequencing to validate nearly 200,000 mutations.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, which is line with the mutation rates generally considered normal in AML.
The area under the precision-recall curve is essentially identical (.995 \& .996) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{E} and Table \ref{table:01}).

\citet{Shi2018} performed multi-region whole exome sequencing on 6 individual breast tumors to a mean target sequencing depth of $160\mathrm{X}$, and validated all variants identified by three different variant calling pipelines.
We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
We estimated a per-base mutation rate for this tumor of $3\cdot10^{-8}$, and the area under the precision-recall curve is identical (.972) for MuTect and \batcave (Figure~\ref{NAR-wes_fig}\textbf{F} and Table \ref{table:01}).
